---
title: "Clinical Drug Trials"
author: "Romaisa Nadeem"
format: pdf
documentclass: article
classoption: [9pt]
header-includes:
  - \usepackage{geometry}
  - \geometry{top=2cm, bottom=2cm, left=2.5cm, right=2.5cm}
  - \usepackage{listings}
  - \lstset{breaklines=true, basicstyle=\ttfamily\small, frame=tb}
---

```{r}
#| echo: false
#| include: false
#| label: setup

library(patchwork)
library(knitr)
library(readr)
library(dplyr)
library(magrittr)
library(tidyr)
library(gt)
library(forcats)
library(stringr)
library(dslabs)
library(ggplot2)
library(viridis)
library(scales)
library(tidyverse)

set.seed(34)
```

## Introduction

Each year, new drugs are approved based on evaluations of their safety and effectiveness through clinical trials. Recently, there has been growing public interest in ensuring fair representation of various subgroups, such as women and racial minorities, in these trials. To address concerns about transparency and representation, the FDA has published Drug Trials Snapshot Summary Report (2015 and 2016) (2017) and the Drugs Trials Snapshots (2023).

I will be exploring two datasets: Clinical Trials: Participant Demographics Dataset (2018) and the Drug Trials Snapshots (2023) dataset.

## About the Dataset

### Clinical Trials: Participant Demographics Dataset

The Clinical Trials: Participant Demographic Data dataset covers demographic details for participants in clinical trials for FDA approved drugs between January 2015 and mid August 2018.

The columns include:

-   **Brand Name**: The commercial name of each drug in the clinical trials.

-   **Drug Indication**: The medical condition the drug is intended to treat.

-   **Percentage of Women**: The proportion of women participants in each trial.

-   **Percentage by Race**: Breakdown of participants by race, including:

    -   **White**
    -   **Black or African American**
    -   **Asian**
    -   **Other**: This category aggregates other races, including American Indian/Alaska Native, Native Hawaiian/Other Pacific Islander, mixed races, unknown, and unreported.

-   **Percentage of Hispanic Ethnicity**: The percentage of Hispanic participants, although this category wasn't included in summary reports for 2015 and 2016.

-   **Percentage Age 65+**: The proportion of trial participants aged 65 and older.

-   **Year**: The year the data was recorded or the drug was approved.

-   The "other rate" column encompasses of any of the following: American Indian/Alaska Native (AI/AN), Native Hawaiian or Other Pacific Islander (NH/OPI), mixed race, multiple races, Unknown, Unreported, and Other.

For drugs approved in 2015 and 2016, percentages for the “Other” category were provided in FDA summary reports. For drugs approved in 2017, ProPublica calculated this percentage by subtracting the sum of other categories from 100 percent. For 2018, ProPublica manually compiled this information for each individual drug.

The categories “Hispanic” and “United States ethnicity” were not included in the FDA’s yearly summary reports for 2015 and 2016, although they sometimes appear in individual drug reports. This category includes one of the following labels: Hispanic, Not Hispanic, and Unknown/Unreported.

### Drug Trials Snapshot Dataset

This dataset is a record of drug trial snapshots from the FDA, covering details about each drug, including its approval information and intended use.

The columns include:

-   **Brand Name (Drug Trials Snapshot)**: Lists the brand names of drugs that have undergone clinical trials.

-   **Drug Name**: The chemical or generic name of each drug.

-   **Original Date of FDA Approval**: The date when the drug received its first FDA approval.

-   **FDA Approved Use on Original Approval Date**: Describes the specific medical condition or purpose for which the drug was approved by the FDA initially.

-   **Brand Name (Prescribing Information)**: Another brand name column, possibly reflecting the name used in prescribing information.

## Questions

In this project, I intend to explore gender and racial biases in clinical drug trials.

I will do show by exploring the following questions:

-   **Are there any potential biases in clinical trials based on age, gender or race?**

-   **Which demographic groups receive the greatest benefit from the drugs being tested in clinical trials?**

While these questions are interesting, they also have very important medical and societal implications. Addressing these biases is essential for promoting equality in healthcare, as unrecognized biases in clinical trials may lead to significant disparities in treatment outcomes, sometimes with severe consequences for patient health and safety. Ensuring fair representation in clinical trials is a step toward more equitable and effective medical treatments for all groups.

## Clinical Trials: Participant Demographics Dataset

## Iteration 1

### Phase 1

1.  Read the csv file into R
    -   Specify column types
2.  Make the column names lowercase
3.  Drop Hispanic and USA columns
    -   Pivot race columns to longer
4.  Pivot Age columns to longer
5.  Identify UIDs
    -   Sort tibble by UID
6.  Display final tibble

### Phase 2

```{r}
# 1. Read in the 1st CSV file. 
clinical_data <- read_csv("./clinical_trials_data.csv",
col_types = "cccccccccccccc",
na = c("N/A", "NR", "Not reported")
)
clinical_data %>%
glimpse()
```

```{r}
# 2. Make column names lowercase
clinical_data %<>% rename_with(tolower)
```

```{r}
# 3.Dropping Hispanic and USA data from 2017 and later
clinical_data %<>%
mutate(`hispanic or latino (2017 and later only)` = NULL, `united states (2017 only)` = NULL)

# Pivot race
clinical_data %<>%
pivot_longer(!c(women, `brand name`, indication, year, notes, `age
65 and older`, `age
75 and older`, `age
80 and older`), names_to = "race", values_to = "percent race")

```

The Hispanic and USA columns are being dropped because they are missing values for more than 50% of the dataset.

```{r}
# 4. Pivot age
clinical_data %<>%
pivot_longer(!c(
  women, `brand name`, indication, year, notes, race,
`percent race`
), names_to = "age", values_to = "percent age")
```

```{r}
clinical_data %>%
glimpse()
```

```{r}
# 5. Identify UID
clinical_data %>%
count(`brand name`, indication, year, race, age) %>%
filter(n > 1) %>%
dim_desc()
```

Looks like a good UID.

```{r}
# Sort tibble by uid
clinical_data %<>% arrange(`brand name`, indication, year, race, age)

clinical_data %>% glimpse()
```

```{r}
# 6. Final tibble
clinical_data %>%
  select(`brand name`, women, year, `percent age`, `percent race`) %>%
  slice_sample(n = 5) %>% 
  kable()

```

## Clinical Trials: Participant Demographics Dataset

## Iteration 2

### Phase 1

1.  Drop notes column
2.  Convert columns to correct data types
    -   year to date type
    -   brand name to factor
    -   race to factor (after renaming "other" values)
    -   age to factor (after renaming "age" values)
3.  Standardize indication column
    -   remove filler words
    -   keep key words and disease names
4.  Remove “%” and “\<” symbols
5.  Convert percentages to numerics
6.  Drop NAs
7.  Display final tibble

### Phase 2

```{r}
# 1. Dropping notes column
clinical_data %<>%
mutate(notes = NULL)
```

```{r}
# 2. Converting data types
clinical_data %<>% 
  # Convert year column to date
  mutate(
    year = make_date(year = year),
    # Convert brand name column to factor
    `brand name` = as_factor(`brand name`),
    # Shorten "all other" data in race column
    race = if_else(str_detect(race, "^all"), "other", race), 
    # Shorten "black or african america" data in race column
    race = if_else(str_detect(race, "^black"), "black", race), 
    # Convert race to factor
    race = as_factor(race)
  )

clinical_data %>%
  glimpse()
```

```{r}
# 2. Continued... Shorten age data and convert to factor
clinical_data %<>%
mutate(
age = case_when(
age == "age\n65 and older" ~ "65 to 74",
age == "age\n75 and older" ~ "75 to 79",
age == "age\n80 and older" ~ "80 and up",
TRUE ~ age
),
age = as_factor(age)
)

clinical_data %>%
glimpse()
```

```{r}
# 3: Standardize indication column, keeping key words
clinical_data <- clinical_data %>%
  mutate(
    indication = case_when(
      str_detect(indication, "[Tt]reatment") ~
        if_else(str_detect(indication, "\\([ˆ()]+\\)"),
                str_c("treatment, ", str_extract(indication, "\\([ˆ()]+\\)")),
                str_to_lower(indication)),
      str_detect(indication, "[Pp]revent") ~
        if_else(str_detect(indication, "\\([ˆ()]+\\)"),
                str_c("preventative treatment, ", str_extract(indication, "\\([ˆ()]+\\)")),
                str_to_lower(indication)),
      TRUE ~ indication
    ),
    # Removing filler words and keeping disease names
    indication = if_else(str_detect(indication, "[Ff]or"), str_remove_all(indication, "[Ff]or"), indication),
    indication = if_else(str_detect(indication, "\\."), str_remove(indication, "\\.$"), indication),
    indication = if_else(str_detect(indication, "associated with"), str_remove_all(indication, "associated with "), indication),
    indication = if_else(str_detect(indication, "adults who"), str_remove_all(indication, "(, adults who).*"), indication),
    indication = if_else(str_detect(indication, "patients two"), str_remove_all(indication, "(, patients two).*"), indication)
  )

# 4: Removing "%" and "<" symbols and converting to numeric
clinical_data <- clinical_data %>%
  mutate(
    women = as.numeric(str_remove_all(women, "[%<]")),
    `percent race` = as.numeric(str_remove_all(`percent race`, "[%<]")),
    `percent age` = as.numeric(str_remove_all(`percent age`, "[%<]"))
  )


clinical_data %>%
glimpse()
```

```{r}
# Check percentages
temp <- clinical_data %>%
pivot_wider(names_from = race, values_from = `percent race`) %>%
rowwise() %>%
mutate(total_percent = sum(c(other, asian, black, white))) %>%
ungroup()
temp %>%
select(`brand name`, other, asian, black, white, total_percent) %>%
filter(total_percent != 100)

```

As seen above, the sum of the race columns does not add up exactly a 100%. This may be due to a error in data collection.

I will leave the data as it is.

```{r}
# 5. Check summary
clinical_data %>%
summary()
```

```{r}
# 5. Drop NAs
clinical_data %<>%
drop_na()

clinical_data %>%
glimpse()
```

There are no outliers in the data, and 865 rows were dropped.

```{r}
# 7. Display final tibble 
clinical_data %>%
slice_sample(n = 5) %>%
kable()
```

## Drug Trials Snapshots Dataset

## Iteration 3

### Phase 1

1.  Read the csv file into R
    -   Skip first column
2.  Change column names to lower
3.  Drop brand name prescribing information column
4.  Identify UID
    -   Sort tibble by UID
5.  Display final tibble

### Phase 2

```{r}
# 1. Read csv
drug_snapshots <- read_csv("./drug_trials_snapshot_fda.csv", skip = 1, show_col_types = FALSE)

drug_snapshots %>% 
  glimpse()

```

```{r}
# 2. Changing column names to lowercase
drug_snapshots %<>% rename_with(tolower)

drug_snapshots %>% 
  glimpse()
```

```{r}
# 3. Drop prescribing name column
drug_snapshots %<>% mutate(
`brand name (prescribing information)` = NULL
)
```

Dropping this column because it does not provide any useful information.

```{r}
# 4. Identify uid
drug_snapshots %>%
count(`original date of fda approval`, `fda approved use on original approval date*`) %>%
filter(n > 1)
```

Seems to be a good UID.

```{r}
# Sort tibble by uid
drug_snapshots %<>% arrange(`original date of fda approval`, `fda approved use on original approval date*`)

# Check structure
drug_snapshots %>% glimpse()
```

```{r}
# 5. Final tibble
drug_snapshots %>%
select(`brand name (drug trials snapshot)`, `original date of fda approval`, `drug name`) %>%
slice_sample(n = 4) %>%
kable()
```

## Drug Trials Snapshots Dataset

## Iteration 4

### Phase 1

1.  Rename columns
2.  Convert approval to date to date type
3.  Standardize approved use column
    -   Keep key words, remove filler words and keep disease names
4.  Drop NAs
5.  Display final tibble

### Phase 2 

```{r}
# 1. Rename columns
drug_snapshots %<>%
rename(
`approved use` = `fda approved use on original approval date*`,
`approval date` = `original date of fda approval`,
`brand name` = `brand name (drug trials snapshot)`
)
drug_snapshots %>%
glimpse()
```

```{r}
# 2. Convert approval date to Date type
drug_snapshots %<>%
mutate(`approval date` = as.Date(`approval date`, format = "%B %d, %Y"))
drug_snapshots %>%
glimpse()
```

```{r}
# 3. Standardize approved use column by keeping key words
drug_snapshots %<>%
  mutate(
    `approved use` = case_when(
      str_detect(`approved use`, "[Tt]reatment") ~
        if_else(
          str_detect(`approved use`, "\\([^)]+\\)"),
          str_c("treatment, ", str_extract(`approved use`, "\\([^)]+\\)")),
          str_to_lower(`approved use`)
        ),
      str_detect(`approved use`, "[Pp]revent") ~
        if_else(
          str_detect(`approved use`, "\\([^)]+\\)"),
          str_c("preventative treatment, ", str_extract(`approved use`, "\\([^)]+\\)")),
          str_to_lower(`approved use`)
        ),
      # Default case to keep `approved use` as it is
      TRUE ~ `approved use`
    )
  )

# Removing filler words and keeping disease names
drug_snapshots %<>%
  mutate(
    `approved use` = str_replace(
      `approved use`, "^[A-Z]", str_to_lower(str_sub(`approved use`, 1, 1))
    ),
    `approved use` = if_else(
      str_detect(`approved use`, "[Ff]or"),
      str_remove_all(`approved use`, "[Ff]or"),
      `approved use`
    ),
    `approved use` = if_else(
      str_detect(`approved use`, "\\."),
      str_remove(`approved use`, "\\.$"),
      `approved use`
    ),
    `approved use` = if_else(
      str_detect(`approved use`, "associated with"),
      str_remove_all(`approved use`, "associated with "),
      `approved use`
    ),
    `approved use` = if_else(
      str_detect(`approved use`, "adults who"),
      str_remove_all(`approved use`, "(?=(, adults who)).*"),
      `approved use`
    ),
    `approved use` = if_else(
      str_detect(`approved use`, "patients two"),
      str_remove_all(`approved use`, "(?=(, patients two)).*"),
      `approved use`
    ),
    # Additional cleanup: Remove leading whitespace and "the " at the start
    `approved use` = str_trim(`approved use`),
    `approved use` = str_replace(`approved use`, "^the ", "")
  )

```

```{r}
drug_snapshots %>% summary()
```

```{r}
drug_snapshots %<>%
drop_na(., `approval date`)
drug_snapshots %>% summary()
```

```{r}
# 5. Final tibble
drug_snapshots %>%
  slice_sample(n = 5) %>%
  kable()
```

## Iteration 5

### Phase 1

1\) Inner join both tibbles

2\) Display final tibble

```{r}
# Join tibbles
clinical_demographics <- inner_join(clinical_data, drug_snapshots, by = "brand name", relationship = "many-to-many")

clinical_demographics %>%
glimpse()
```

For the purposes of this project, I will be focusing on the clinical trials data to derive insights on clinical participant demographics. The drug snapshots adds more information to the initial table.

```{r}
# Final tibble
clinical_demographics %>%
slice_sample(n = 3) %>%
kable()
```

## Results and Visualization 

```{r plot_1, fig.width=7, fig.height=5}
p1 <- clinical_demographics %>%  
  ggplot(aes(x = str_to_title(race), y = `percent race`)) +  
  geom_violin(aes(fill = race, colour = race)) +  
  geom_boxplot(width = 0.1) +  
  geom_jitter(alpha = 0.1) +  
  labs(  
    x = "Race",   
    y = "Percent of People from Race"  
  ) +  
  scale_color_viridis_d() +  
  theme(legend.position = "none", plot.caption = element_text(hjust = 0))  

p1 + facet_wrap(. ~ age, labeller = label_both) +  
  guides(x = guide_axis(angle = 45)) +  
  plot_annotation(title = "Graph 1")
```

Graph 1 illustrates the distribution of clinical trial participants across different racial groups and age categories using box and violin plots.

### **Age Groups:**

-   **65 to 74**:

    -   The "White" category shows a wide distribution, indicating variability in the percentage of participants.

    -   Other racial groups like "Asian," "Black," and "Other" have narrower distributions, suggesting less variability.

-   **75 to 79**:

    -   The "White" category continues to show a wider spread, though less pronounced than the 65 to 74 group.

    -   Other racial groups maintain narrower distributions.

-   **80 and up**:

    -   The "White" category again shows a wider distribution, indicating variability.

    -   Other racial groups remain consistent with narrower spreads.

### **Racial Distribution:**

-   The majority of participants identify as "White," with a significant density of points indicating a higher percentage.

-   Other racial groups have fewer participants, with distributions centered around the mean and fewer outliers.

### **Overall Trends:**

-   The "White" category shows more variation across age groups, not centered around the mean.

-   Other racial groups show less variation, with percentages more centered around the mean and some outliers.

```{r plot_2, fig.width=10, fig.height=5}
p2 <- clinical_demographics %>%  
  ggplot(aes(x = women)) +  
  stat_ecdf() +  
  geom_vline(aes(xintercept = 50), colour = "red") +  
  labs(x = "Percentage of Women Patients in the Trial", y = "Cumulative Frequency") +  
  facet_grid(age ~ str_to_title(race)) +  
  guides(x = guide_axis(angle = 45)) +  
  theme(axis.text.x = element_text(color = "black"), legend.position = "none")  

p3 <- clinical_demographics %>%  
  mutate(specific_values = FALSE) %>%  
  bind_rows(data.frame(x = 50, specific_values = TRUE)) %>%  
  mutate(y = ecdf(women)(x)) %>%  
  ggplot(aes(x = women)) +  
  stat_ecdf(data = . %>% filter(!specific_values), aes(colour = race)) +  
  labs(x = "Percentage of Women Patients in the Trial", y = "Cumulative Frequency") +  
  geom_segment(data = . %>%  
    filter(specific_values), aes(y = y, x = 0, xend = x, yend = y), color = "red") +  
  geom_segment(data = . %>%  
    filter(specific_values), aes(y = 0, x = x, xend = x, yend = y), color = "red") +  
  scale_color_viridis_d()  

# Combine plots  
combined_plot <- p2 + p3 +  
  plot_annotation(  
    title = "Graph 2",   
    tag_levels = "1",   
    tag_prefix = "p"  
  ) +  
  theme(plot.caption = element_text(hjust = 1))  

# Print the combined plot  
print(combined_plot)
```

P1 and p2 in graph 2 provide insights into the distribution of women participants in clinical trials, segmented by race and age groups.

### **P1:**

-   **Cumulative Frequency by Race and Age**:

    -   This panel displays cumulative frequency plots for different racial groups (Asian, Black, Other, White) across three age categories: 65 to 74, 75 to 79, and 80 and up.

    -   The overlapping distributions indicate that the percentage of women participants is similar across racial groups, as the curves for each race largely coincide.

    -   However, there is a noticeable variation in the distribution of women participants across age groups. Younger age groups show a lower frequency of trials with 50% women participants compared to older age groups, suggesting that older participants are more likely to be represented in trials.

### **P2:**

-   **Overall Cumulative Frequency**:

    -   This plot shows the cumulative frequency of women participants across all trials, with a focus on the percentage of women.

    -   The red vertical line at 50% indicates a threshold where the frequency of trials with at least 50% women participants is highlighted.

    -   The cumulative distribution suggests that while there are trials with a higher percentage of women, the overall representation is still skewed, particularly in younger age groups.

### **Key Observations:**

1.  **Representation of Women**:

    -   There is a larger representation of women in clinical trials for older age groups compared to younger ones, as indicated by the cumulative frequency plots.

2.  **Racial Distribution**:

    -   Across all age groups, the predominant race in clinical trials is White/Caucasian, with similar distribution patterns among different races, although this varies by age.

3.  **Trial Participation**:

    -   Despite a larger number of clinical trials involving younger participants, the percentage of women in these trials is relatively low, highlighting a disparity in representation.

## Discussion

The analysis of clinical trial demographics reveals a clear pattern regarding which population benefits most from drug trials and subsequent drug development. Based on the comprehensive data examination, White males between the ages of 65 and 74 emerge as the predominantly represented demographic in clinical trials. This overrepresentation creates a significant skew in drug development and testing, resulting in medications that are primarily validated for this specific demographic group.

The implications of this demographic bias are far-reaching and problematic on multiple levels. When clinical trials consistently favor one demographic group, the resulting pharmaceutical developments are inherently more tailored to that population's physiological characteristics and needs. This means that White males in their mid-to-late 60s and early 70s receive medications that have been extensively tested on individuals sharing their demographic profile, potentially leading to more predictable and effective treatment outcomes for this group.

However, this bias creates substantial medical concerns for other demographic groups. The human body's response to medications can vary significantly based on factors such as age, gender, racial background, and genetic makeup. When drugs are primarily tested on a narrow demographic segment, their effectiveness and potential side effects on other populations remain largely unknown until the medication enters widespread use. This creates a dangerous scenario where medications approved based on positive results in one demographic group may produce unexpected or potentially harmful effects in others.

The medical implications of this bias extend beyond simple effectiveness concerns. Different populations may require different dosages, experience varying side effects, or respond differently to treatment protocols. Without adequate representation in clinical trials, these variations remain undiscovered during the testing phase, potentially endangering lives when the drugs are prescribed to members of underrepresented groups.

This situation highlights the urgent need for reform in clinical trial recruitment and participation. To ensure safe and effective medical treatments for all populations, clinical trials must actively seek and include participants from diverse backgrounds, ages, and genders. This diversity in testing would lead to more comprehensive understanding of drug effects across different demographic groups and ultimately result in safer, more effective medical treatments for the entire population.

The current bias in clinical trials not only raises ethical concerns about equality in medical research but also presents a significant challenge to the fundamental goal of medicine: to provide safe and effective treatment for all patients, regardless of their demographic background. Moving forward, the medical research community must prioritize the development of more inclusive clinical trial protocols that ensure representation across all demographic groups, thereby working toward true equality in medicine and healthcare outcomes.

## Conclusion 

In conclusion, the analysis of clinical trial demographics underscores a significant bias favoring White males aged 65 to 74, leading to skewed drug development outcomes that may not effectively serve other demographic groups. This over representation raises ethical and medical concerns, as medications validated primarily on this narrow population may pose risks or prove ineffective for women, younger individuals, and racial minorities. To address these disparities, it is crucial for the medical research community to implement more inclusive recruitment strategies that ensure diverse representation in clinical trials. By doing so, we can enhance the safety and efficacy of treatments for all patients, ultimately advancing equity in healthcare and improving health outcomes across the population.

## References

*Clinical Trials: Participant Demographic Data*. (n.d.). ProPublica Data Store. https://www.propublica.org/datastore/dataset/cancer-clinical-trials-demographics-data

FDA. (2019). *Drug Trials Snapshots*. U.S. Food and Drug Administration. https://www.fda.gov/drugs/drug-approvals-and-databases/drug-trials-snapshots

FDA. (2024). *Drug Trials Snapshots Summary Report (2015 and 2016)*. U.S. Food and Drug Administration. https://www.fda.gov/drugs/drug-approvals-and-databases/drug-trials-snapshots-summary-report-2015-and-2016
